Modeling Second-Language Learning from a Psychological Perspective

### 2018-04-12 doc/naaclhlt2018-latex/paper.tex
addressed reviews, comments welcome

\hypersetup{draft}
\newcommand\BibTeX{B{\sc ib}\TeX}

@@

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of learning in large samples. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousands of users completing millions of exercises to 
predict patterns of future translation mistakes in held-out data.  The data was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team which ended in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

@@

In the training data, all instances are labeled as correct or incorrect, so the
label for the previous encounter is always available. In the test data, labels
are unavailable, so predictions must be made using a mix of labeled and
unlabeled past encounters. To generate training-set features that are 
comparable to test-set features, we selectively ignored some labels when encoding temporal features on
the training set. Specifically, for each user we first calculated the number of
exercises $n$ in the true test set. Then, when encoding the features for each
instance, we selected a random integer $r$ in the range $[1,n]$, and ignored labels
in the prior $r$ exercises. That is, we encode features for the current instance
as though other instances in those prior exercises were unlabeled, and ignore
updates to the error averages from those exercises. The result of this process
is that each instance in the training set is encoded as though it were between
one and $n$ exercises into the test set.

@@

\begin{table}[t!]
\small
\begin{center}
\begin{tabular}{|l|llll|}
  % \hline   & \multicolumn{3}{l}{\bf Value} & \\
  \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
  num\_leaves & 256 & 512 & 512 & 1024 \\
  learning\_rate & .05 & .05 & .05 & .05 \\
  min\_data\_in\_leaf & 100 & 100 & 100 & 100 \\
  num\_boost\_rounds & 750 & 650 & 600 & 750 \\
  cat\_smooth & 200 & 200 & 200& 200 \\
  feature\_fraction & .7 & .7 & .7 & .7 \\
  max\_cat\_threshold & 32 & 32 & 32& 64 \\
\hline
\end{tabular}
\end{center}
\caption{\label{lightgbm-params} Parameters of final LightGBM models. See
  LightGBM documentation for more information; all other parameters were left at
their default values.}
\end{table}

@@

We trained a model for each of the three language tracks of {\tt en\_es}, {\tt es\_en},
and {\tt fr\_en}, and also trained a model on the combined data from
all three tracks, adding an additional ``language'' feature. Following model
training, we averaged the predictions of each single-language model with that of
the all-language model to form our final predictions.

@@

The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
  es\_en}, and $.8540$ on {\tt fr\_en}. For reference this placed us within .01 of the winning
  entry for each problem ($.8613$ on {\tt en\_es}, $.8383$ on {\tt
  es\_en}, and $.8570$ on {\tt fr\_en}).  Also note that the Duolingo-provided baseline
  model (L2-regularized regression trained with stochastic gradient descent weighted
  by frequency) obtains $.7737$ on {\tt en\_es}, $.7456$ on {\tt
  es\_en}, and $.7707$ on {\tt fr\_en}.
  We did not attempt to optimize the model's
F1 score, but the F1 score could likely be improved (at the cost of
increased log loss) by finding the
rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.

@@

To better understand which features or groups of features were most important to
our model's predictions, we conducted a set of experiments in which we lesioned
(i.e., removed) a group of features and re-trained the model on the {\tt train}
set, evaluating performance on the {\tt dev} set. For simplicity, we ran each of the
lesioned models on all language data and report the average
performance. We did not run individual-language models as we did for our primary model.
The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
models are as follows.
\begin{description}
  \item[none:] All features are included.  
  \item[user:] All user-level features, including the user ID and other
    calculated features like entropy and measures of exercise bursts, are
    removed.
  \item[userid \& user other:] Only user ID or only the calculated
    features user features, respectively, are removed.
  \item[word:] Token and token root IDs; previous, next, and dependency-tree
    root word IDs; and morphological, part of speech, and dependency tree
    features are removed. This does not include {\bf external} features listed below.
  \item[word id \& word other:] Only word IDs or only other
    word features, respectively, are removed. {\bf word other} does not include {\bf external} features listed below.
  \item[neighbors:] Both word IDs and other word features are removed, but
    only for the previous, next, and dependency-tree root words. Information
    about the present word is maintained.
  \item[external:] External information about the word, including corpus
    frequency, Levenshtein distance from translation, and age of acquisition,
    are removed.
  \item[temporal:] Temporal information, including number and timing of
    past encounters with the word and error tracking information, is removed.
\end{description}

@@

Though not plotted above, we also ran a model lesioning exercise-level features
including client, session type, format, and exercise duration. This model
achieved an AUROC of $.787$, far lower than any other lesion. This points to the
fact that \emph{how} memory is assessed often includes observed performance 
\cite[e.g., the large literature in psychology on the difference between recall and recognition memory,][]{yonelinas2002nature}.

>>>>

% \hypersetup{draft}
\newcommand\BibTeX{B{\sc ib}\TeX}

@@

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of learning in large samples. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousands of users completing millions of exercises to 
predict patterns of future translation mistakes in held-out data.  The data was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team, which finished in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

@@

In the training data, all instances are labeled as correct or incorrect, so the
label for the previous encounter is always available. In the test data, labels
are unavailable, so predictions must be made using a mix of labeled and
unlabeled past encounters. In particular, for a user's test set with $n$
exercises, each exercise will have between zero and $n-1$ preceding unlabeled exercises.

To generate training-set features that are 
comparable to test-set features, we selectively ignored some labels when encoding temporal features on
the training set. Specifically, for each user we first calculated the number of
exercises $n$ in their true test set\footnote{If the size of the test set were
  not available, it could be estimated based on the fact that it is
  approximately 5\% of each participant's data.}. Then, when encoding the features for each
training instance, we selected a random integer $r$ in the range $[0,n-1]$, and
ignored the labels in the prior $r$ exercises. That is, we encoded features for the instance
as though other instances in those prior exercises were unlabeled, and ignored
updates to the error averages from those exercises. The result of this process
is that each instance in the training set was encoded as though it were between
one and $n$ exercises into the test set.

@@

\begin{table}[t!]
\small
\begin{center}
\begin{tabular}{|l|llll|}
  % \hline   & \multicolumn{3}{l}{\bf Value} & \\
  \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
  {\tt num\_leaves} & 256 & 512 & 512 & 1024 \\
  {\tt learning\_rate} & .05 & .05 & .05 & .05 \\
  {\tt min\_data\_in\_leaf} & 100 & 100 & 100 & 100 \\
  {\tt num\_boost\_rounds} & 750 & 650 & 600 & 750 \\
  {\tt cat\_smooth} & 200 & 200 & 200& 200 \\
  {\tt feature\_fraction} & .7 & .7 & .7 & .7 \\
  {\tt max\_cat\_threshold} & 32 & 32 & 32& 64 \\
\hline
\end{tabular}
\end{center}
\caption{\label{lightgbm-params} Parameters of final LightGBM models. See
  LightGBM documentation for more information; all other parameters were left at
their default values.}
\end{table}

@@

We trained a model for each of the three language tracks of {\tt en\_es}, {\tt es\_en},
and {\tt fr\_en}, and also trained a model on the combined data from
all three tracks, adding an additional ``language'' feature. Following model
training, we averaged the predictions of each single-language model with that of
the all-language model to form our final predictions. Informal experimentation
showed that model averaging provided a modest performance boost, and that
weighted averages did not clearly outperform a simple average.

@@

The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
  es\_en}, and $.8540$ on {\tt fr\_en}. For reference this placed us within .01 of the winning
  entry for each problem ($.8613$ on {\tt en\_es}, $.8383$ on {\tt
  es\_en}, and $.8570$ on {\tt fr\_en}).  Also note that the Duolingo-provided baseline
  model (L2-regularized regression trained with stochastic gradient descent weighted
  by frequency) obtains $.7737$ on {\tt en\_es}, $.7456$ on {\tt
  es\_en}, and $.7707$ on {\tt fr\_en}.
We did not attempt to optimize F1 score, the competition's secondary evaluation metric.
%   We did not attempt to optimize the model's
% F1 score., but the F1 score could likely be improved (at the cost of
% increased log loss) by finding the
% rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
% the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.

@@

To better understand which features or groups of features were most important to
our model's predictions, we conducted a set of experiments in which we lesioned
(i.e., removed) a group of features and re-trained the model on the {\tt train}
set, evaluating performance on the {\tt dev} set. For simplicity, we ran each of the
lesioned models on all language data and report the average
performance. We did not run individual-language models as we did for our primary model.
The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
models are as follows.
\begin{description}
  \item[none:] All features are included.  
  \item[user all:] All user-level features, including the user ID and other
    calculated features like entropy and measures of exercise bursts, are
    removed.
  \item[user id \& user other:] Only user ID or only the calculated
    user features, respectively, are removed.
  \item[word all:] Token and token root IDs; previous, next, and dependency-tree
    root word IDs; and morphological, part of speech, and dependency tree
    features are removed. This does not include {\bf external} features listed below.
  \item[word id \& word other:] Only word IDs or only other
    word features, respectively, are removed. {\bf word other} does not include {\bf external} features listed below.
  \item[neighbors:] Both word IDs and other word features are removed, but
    only for the previous, next, and dependency-tree root words. Information
    about the present word is maintained.
  \item[external:] External information about the word, including corpus
    frequency, Levenshtein distance from translation, and age of acquisition,
    are removed.
  \item[temporal:] Temporal information, including number and timing of
    past encounters with the word and error tracking information, is removed.
\end{description}

@@

Though not plotted above, we also ran a model lesioning exercise-level features
including client, session type, format, and exercise duration. This model
achieved an AUROC of $.787$, far lower than any other lesion. This points to the
fact that the manner in which memory is assessed often affects observed performance 
\cite[e.g., the large literature in psychology on the difference between recall and recognition memory,][]{yonelinas2002nature}.
